---
title: "External Data"
page-navigation: false
format:
  html: 
    toc: false
webr:
  packages: 'data.tree'
---

```{r, echo = FALSE, include = FALSE}
source(here::here("R", "buttons.R"))
```

```{webr-r}
#| autorun: true
#| context: setup

dir.create("code")
dir.create("data")
dir.create("output")
dir.create("figures")

download.file("https://openstatprog.com/raw-data/boys.dat", "data/boys.dat")
download.file("https://openstatprog.com/raw-data/boys.RData", "data/boys.RData")
download.file("https://openstatprog.com/raw-data/boys_eu.csv", "data/boys_eu.csv")

download.file("https://openstatprog.com/raw-data/diabetes.txt", "data/diabetes.txt")
download.file("https://openstatprog.com/raw-data/diabetes.rds", "data/diabetes.rds")

download.file("https://openstatprog.com/raw-data/iris.txt", "data/iris.txt")
download.file("https://openstatprog.com/raw-data/iris.csv", "data/iris.csv")

download.file("https://openstatprog.com/raw-data/titanic.rds", "data/titanic.rds")
download.file("https://openstatprog.com/raw-data/utmb_2017.csv", "data/utmb_2017.csv")
download.file("https://openstatprog.com/raw-data/mtcars.sav", "data/mtcars.sav")
download.file("https://openstatprog.com/raw-data/starwars.sav", "data/starwars.sav")
download.file("https://openstatprog.com/raw-data/example_data.xlsx", "data/example_data.xlsx")

files <- list.files("./", recursive = TRUE, include.dirs = TRUE, full.names = TRUE)
dir_tree <- data.tree::as.Node(data.frame(pathString = files))
```

Welcome to the *External Data* module. In this module you will learn how to use external datasets in your R analyses.
We will first cover methods for loading datasets stored in the most common formats.

## Learning Goals

After completing this module, you will be able to load data stored as:

1. Delimited text files (e.g., tab-delimited, CSV)
1. R's native data files
1. SPSS data files
1. Excel workbooks

## Directory Structure

This module is all about reading data files from your computer's hard drive. That focus presents a bit of a technical
problem. The *webr* sessions in which you complete the exercises for these tutorials runs directly in your browser and
cannot access your computer's actual file system. So, we've created a synthetic directory structure for you to use in
this module's practice problems. The following dendrogram shows the contents of the working directory for the *webr*
sessions that you'll be using in this module. Note the data files stored in the `./data/` directory: those are the files
we'll be loading in this module.

```{webr-r}
#| autorun: true
#| context: output

print(dir_tree)
```

------------------------------------------------------------------------------------------------------------------------

Click the button below to get started.

```{r}
#| echo: false
#| results: asis
begin_button("/tutorials/loading/index.html")
```
