---
title: "Native Data Formats"
webr:
  packages: ['data.tree', 'here']
---

```{webr-r}
#| autorun: true
#| context: setup

dir.create("code")
dir.create("data")
dir.create("output")
dir.create("figures")

download.file("https://openstatprog.com/raw-data/boys.dat", "data/boys.dat")
download.file("https://openstatprog.com/raw-data/boys.RData", "data/boys.RData")
download.file("https://openstatprog.com/raw-data/boys_eu.csv", "data/boys_eu.csv")

download.file("https://openstatprog.com/raw-data/diabetes.txt", "data/diabetes.txt")
download.file("https://openstatprog.com/raw-data/diabetes.rds", "data/diabetes.rds")

download.file("https://openstatprog.com/raw-data/iris.txt", "data/iris.txt")
download.file("https://openstatprog.com/raw-data/iris.csv", "data/iris.csv")

download.file("https://openstatprog.com/raw-data/titanic.rds", "data/titanic.rds")
download.file("https://openstatprog.com/raw-data/utmb_2017.csv", "data/utmb_2017.csv")
download.file("https://openstatprog.com/raw-data/mtcars.sav", "data/mtcars.sav")
download.file("https://openstatprog.com/raw-data/starwars.sav", "data/starwars.sav")
download.file("https://openstatprog.com/raw-data/example_data.xlsx", "data/example_data.xlsx")

files    <- list.files("./", recursive = TRUE, include.dirs = TRUE, full.names = TRUE)
dir_tree <- data.tree::as.Node(data.frame(pathString = files))
```

R provides two native data formats: `RData` and `RDS`.

## RData Files

Although we can certain store datasets as RData files, it's a bit misleading to characterize the RData format as a tool
for storing datasets. RData files are actually compressed snapshots of an entire R environment. So, when we "load" an RData
file, we're not really loading a single dataset; rather, we're "restoring" the contents of the environment stored in the
file. Of course, the environment we restore may contain only a single datasets, in which case the end result will be
functional equivalent to loading that dataset. It's worth remembering, though, that the RData format is a more general
tool than we need for this job.

We use the `load()` function to load RData files. In the following code chunk, we load the `"boys.RData"` object. This
file contains only a single data frame called `boys`.

```{r}
dataDir <- "data"

# List the contents of the current environment
ls()

# Load the objects stored in 'boys.RData'
load(here::here(dataDir, "boys.RData"))

# Now we have a new data frame in our environment
ls()
head(boys)
```

Notice that we don't assign the return value of `load()` to an object. When we run `load()`, we simply add all the
objects stored therein to our current environment. Hence, RData files are best suited to capturing an instantaneous
*snapshot* of your current R session so that you can exactly restore the current state of your environment sometime in
the future.

In the following code chunk, we use an RData object to restore the snapshot of a previous R session.

```{r}
# The environment contains only the objects we loaded in the last example
ls()

# Load the objects stored in 'snapshot.RData'
load(here::here(dataDir, "snapshot.RData"))

# Now, we've added several new objects to our session
ls()
```

After we restore the objects stored in "snapshot.RData", we add two new datasets, `iris` and `attitude`, but we also
restore a set of regression model results, `out`, and a ggplot object, `p1`.

```{r}
str(iris)
str(attitude)
summary(out)
print(p1)
```

We could now pick up with our previous analysis exactly where we left off when we created "snapshot.RData".

## RDS Files

RData files are much better for the kind of "back-up and restore" operations that we see in the preceding example than
they are for storing individual datasets. Fortunately, R has a second native data format that is ideally suited for
storing datasets. RDS files are specifically designed to store single R objects like the data frames we typically use to
hold datasets. We use the `readRDS()` function to load RDS files.

```{r}
titanic <- readRDS(here::here(dataDir, "titanic.rds"))
str(titanic)
```

Notice how the `readRDS()` call follows the same conventions as the other data-ingest functions we've covered. Most
importantly, `readRDS()` returns a single R object, and we assign that object to a variable in our environment.

::: practice

1. Load the dataset saved as "./data/diabetes.rds".
1. Use the `str()` function to compare the structure of the dataset you loaded to the `diabetes0` object loaded below.

Are there any differences between these two objects? If so, what are the differences?

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
diabetes0 <- read.table(here::here("data", "diabetes.txt"),
                        header = TRUE,
                        sep = "\t")
```

## {{< var tab_title.dir >}}

The following dendrogram illustrates the structure of the working directory for this *webr* session.

```{webr-r}
#| autorun: true
#| context: output

print(dir_tree)
```

## {{< var tab_title.solution >}}

```{r}
diabetes0 <- read.table(here::here("data", "diabetes.txt"), header = TRUE, sep = "\t")
diabetes1 <- readRDS(here::here("data", "diabetes.rds"))
str(diabetes0)
str(diabetes1)
```

The main difference between these two objects lies in the fact that `diabetes1` preserves all the R-specific formatting.
Notably, the `sex` variable is stored as a two-level factor in `diabetes1`. In `diabetes0`, on the other hand, `sex` is
a character vector. 

::::

:::
